---
title: "Introduction to the iimi package"
author: "Haochen Ning"
date: '`r paste("First created on 2023-04-29. Updated on", Sys.Date())`'
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to the iimi package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES/reference.bib
---

```{r, echo=FALSE, warning=FALSE}
library(httr)
```


## 1. Introduction

This vignette aims to give an introduction on how to use the `iimi` package for
plant virus diagnostics and how to visualize the coverage profile for the 
sample mapping. We also included a tutorial on creating unreliable regions.

### 1.1. Installation

First, let's install necessary packages. You may skip this step if you have 
installed the packages before.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# install Biostrings
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("GenomicAlignments", "Rsamtools"))
BiocManager::install("Biostrings")

# install iimi
install.packages(c("iimi", "dplyr"))
```

### 1.2. Loading packages

We will load necessary packages before we start any analysis.

```{r, warning=FALSE, message=FALSE}
library(iimi)

library(Biostrings)

library(dplyr)
```

### 1.3. Data Pre-processing

We provide three example coverage profiles to demonstrate how to use `iimi`. 
These files are sourced from the dataset used in the VirHunter paper [@virhunter],
which we also utilized for external validation in our manuscript [@iimi]. You can
download these files directly from the Recherche Data Gouv website [@sugarbeet]. 
We recommend storing all BAM files in a single folder for ease of access.

To get started creating coverage profiles and feature-extracted data frames, the
process involves four steps: 
(1) Downloading the raw data in FASTQ format from the link above
(2) Use Bowtie2 to map the FASTQ files (paired-end or single-end) against the [official Virtool virus data base (ver. 1.4.0)](https://github.com/virtool/ref-plant-viruses) to get SAM outputs
(3) Convert SAM to indexed and sorted BAM using Samtools
(4) Generate coverage profiles and feature-extracted data frames using `iimi`
functions (tutorials in the next section)

We recommend Bowtie2 or minimap2 since we have tried both and they yield similar
result with minimap2 having a slight decrease. We let both software to report all
alignments (`-a` mode for Bowtie2, `--secondary=yes` for minimap2). You can also 
use other mapping tools.

Next, we provide a short tutorial to guide you through using `iimi` functions to
make predictions on the real data.


## 2. Converting BAM file(s) into coverage profiles and feature-extracted data frame

Let's convert the indexed and sorted BAM file(s) into coverage profiles and 
feature-extracted data frame from the previous section.

We will use the coverage profiles to visualize the mapping information. The
feature-extracted data frame will be used in the model training and testing 
process.

**Note that both training and testing data need to go through the conversion step.**

In our example, we stored the conversion for both the testing and training 
datasets in the same object. You can do the conversion separately for your data.

**Important: the example code does not work unless the path to the folder that stores your BAM files is provided.**

### 2.1. State the path to the folder of your BAM files

If you already have coverage profiles in run-length encoding (RLE) format, go 
to section 2.2.

```{r, eval=FALSE, warning=FALSE}
path_to_bamfiles <- list.files(
  path = "path/to/your/BAM/files/folder",
  pattern = "bam$",
  full.names = TRUE,
  include.dirs = TRUE
)
```

### 2.2. Create a data frame that contains the coverage profiles.

#### 2.2.1. Convert BAM files to a list of RLEs.

You may skip this step if you already have converted them to RLE format.

```{r, eval=FALSE, warning=FALSE}
example_cov <- convert_bam_to_rle(bam_file = "path_to_bamfiles")
```

#### 2.2.2. Convert the list of RLEs to a feature-extracted data frame.

This section explains how to convert the RLE format to feature-extracted 
dataframes, with options for handling unreliable regions.

By default, the package uses provided unreliable regions and enables mappability
profiling and filtering to eliminate false peaks. This is the recommended 
approach for most users. 
* To use the default settings, no additional code is required.
* To disable this feature, set `unreliable_region_enabled = FALSE` when calling `convert_rle_to_df()`.

If you wish to use your own unreliable regions, refer to section 3.3 for instructions on creating custom unreliable regions and input your custom unreliable regions using the `unreliable_region_df` parameter in `convert_rle_to_df()`.

**Note: Customization of unreliable regions must be done before calling `convert_rle_to_df()`. Most users can skip this step and use the provided unreliable regions.**

```{r, eval=FALSE, warning=FALSE}
# Using default settings (recommended)
df <- convert_rle_to_df(example_cov)

# Disabling unreliable region processing
df <-
  convert_rle_to_df(example_cov, unreliable_region_enabled = FALSE)

# Using custom unreliable regions
# Refer to section 3.3 for details
custom_regions <- create_custom_unreliable_regions()
df <-
  convert_rle_to_df(example_cov, unreliable_region_df = custom_regions)
```


## 3. Predicting the plant sample(s)

To make predictions, use the converted mapping result of the sample(s) that you 
wish to detect as the input, `newdata`. Make sure you have converted the indexed
and sorted BAM files into feature-extracted data frame from the section above.

After preparing your test sample, you can choose to test the data using our
provided training model or the model you trained using `train_iimi()`. The 
tutorial of training your own model is provided in the next section.

**Note: if you wish to customize unreliable regions, please go to 3.3.**

### 3.1 Using pre-trained models and no customization

If you wish to use provided training model, only input your data to `newdata` 
and choose a method of your wish using `predict_iimi()`.

There are three methods that you may choose from: `xgb`, `en`, and `rf`, which
stand for pre-trained XGBoost, elastic net, and random forest models. The 
example below uses the pre-trained XGBoost model.

```{r, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
prediction_default <- predict_iimi(newdata = df, method = "xgb")
```

The detection of your plant sample(s) is finished. The prediction is `TRUE` if 
virus infected the sample, `FALSE` if virus did not infect the sample.


### 3.2. Customizing your own model

If you would like to train your own model and use this model to test your data, 
you can use the codes below to train a new model with your own data.

Ideally, the number of the samples used to train the model should be bigger than
100. However, we are only providing a tutorial on how to use the 
`train_iimi()` function, only two samples are used to train the model since 
`example_cov()` only contains three in-house data's coverage information. 

Now, we need to prepare our training data. We are using a 80/20 random split 
to split the three samples. This means that two samples are used as the training
data, and one sample is used as the testing data. If you are training your own 
data, the training data is the data that you want to train the model on; the
testing data is the data that you would like to have a prediction on.

Here are some definitions/explanation of the objects to input in `train_iimi()`:

1. `train_x`: the feature-extracted data frame of plant samples that you would 
like to train `iimi` model on. Make sure that you have mapped the samples to the
virus database and converted the mapping result to sorted and indexes BAM files.

2. `train_y`: the known truth or labels for your `train_x` data. Please label
the data to make sure that it has a detection label for virus segments as well.

3. `test_x`: the feature-extracted data frame of plant samples that you would 
like to predict using your trained `iimi` model. Make sure that you have mapped 
the samples to the virus database and converted the mapping result to sorted 
and indexes BAM files.

```{r, eval=FALSE}
# preparing the train and test data

# spliting into 80-20 train and test data set with the three plant samples
set.seed(123)
train_names <- sample(levels(as.factor(df$sample_id)),
                      length(unique(df$sample_id)) * 0.8)

# trian data
train_x = df[df$sample_id %in% train_names,]
# test data
test_x = df[df$sample_id %in% train_names == F,]

# preparing labels
train_y = c()
for (ii in 1:nrow(train_x)) {
  train_y = append(train_y, example_diag[train_x$seg_id[ii],
                                         train_x$sample_id[ii]])
}
```

Then, we plug in the variables into the `train_iimi` function with the default
XGBoost model to train your custom model.

```{r, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
fit <- train_iimi(train_x = train_x, train_y = train_y)
```

Now, we have a trained model using the toy data.

Then, the process to detect which viruses infect the plant sample(s) is the 
same as previously described, except we are using a customized trained model.

```{r, eval=FALSE}
prediction_customized <-
  predict_iimi(newdata = test_x,
               trained_model = fit)
```

The detection of the plant sample(s) is finished. The interpretation is the same
as above.


### 3.3. Customizing unreliable regions

**Note: if you would like to create your own unreliable regions, please customize them first, then extract features to build a data frame from section 2.2.2. using customized unreliable regions.**

If you would like to create your own unreliable regions besides from using your 
own training model, you may use `create_mappability_profile()` and
`high_nucleotide_regions()`. Both functions' output is a data frame with the 
start and end position of the unmappable region, the virus that the region is 
on, and the category that it belongs to.

```{r}
# An example of the provided unreliable regions
unreliable_regions %>% group_by(Categories) %>% sample_n(2)
```

Unreliable regions contains (1) mappability profile (2) high nucleotide content
regions.

* Mappability profile is a profile of areas on a virus genome that can be mapped 
to (1) other viruses or (2) host genome. We choose Arabidopsis Thaliana as our
host genome.

* High nucleotide content regions is a profile of areas on a virus genome that 
has (1) high GC content and/or (2) high A nucleotide percentage.

Including these two profiles into `iimi` ensures that there are no false peaks 
like the ones described in the previous section.

#### 3.3.1. Mappability profile

Here is a short tutorial to make mappability profile.

1. Split each of the virus segment from the virus database into a sliding 
window series with window size of your choice and with step size 1. The default
value for window size is 75. You may choose any window size you want.

2. Map one virus segment with each other, until you finish mapping it to all
virus segments in the virus database. Also map the virus segment with a host 
genome of your choice. We chose to use Arabidopsis Thaliana.

3. Sort and index the resulted BAM files from the mapping step.

4. Assemble the mappability profile:

```{r, eval=FALSE}
# if you would like to keep unmappable regions that can be mapped to other
# viruses or the host genome separate into two data frames, you may use the
# following code:

# input your own path that you would want to store regions on a virus that can
# be mapped to another virus
# you can customize the name of this type of mappability profile
mappability_profile_virus <-
  create_mappability_profile("path/to/bam/files/folder/virus", category = "Unmappable region (virus)")

# input your own path that you would want to store regions on a virus that can
# be mapped to the host genome
# you can customize the name of this type of mappability profile
mappability_profile_host <-
  create_mappability_profile("path/to/bam/files/folder/host", category = "Unmappable region (host)")

# if you would like to keep everything in the same data frame, you may use the
# following code:
mappability_profile <-
  create_mappability_profile("path/to/bam/files/folder/of/both/types/", category = "Unmappable region")
```

#### 3.3.2. High nucleotide content regions

Creating the high nucleotide content regions is much easier than the mappability
profile. We only need to use `create_high_nucleotide_content()` function.

Here is an example:

```{r, eval=FALSE}
high_nucleotide_regions <-
  create_high_nucleotide_content(gc = 0.6, a = 0.45)
```

The default threshold for GC content is 60% and is 45% for A%. The thresholds 
are customizable.

Now, we can combine these two regions into the final unreliable regions. 
You can use them to convert your training and testing data to feature-extracted data frames. Please refer to section 2.2.2. to see how to do so.


## 4. Visualizing the coverage profiles

Next, we can visualize the coverage profile by using the `plot_cov()` function.

* `plot_cov()`: plots the coverage profile of the plant sample and the percentage
  of A nucleotides and GC content for a sliding window of k-mer with the step as 
  1. We used the default setting of k = 75.

```{r, fig.width=7, fig.height=5}

oldpar <- par(mfrow = c(1, 2))

## if you wish to plot all segments of one sample, you can try:
# plot_cov(covs = example_cov["S1"])

## if you wish to plot all segments from all samples, you can try:
# plot_cov(covs = example_cov)

## if you wish to plot certain segments from one sample, you can try:
segs = c("42jtlrir", "m0kacxse")
covs_selected = list()
covs_selected$`305S` <-
  example_cov$`305S`[segs]

## if you have many segments that you would want to plot, you can try the following code with the numbers changed
## to find the index of your desired segments:

covs_selected$S1 <-
  example_cov$S1[names(example_cov$S1)[c(1,72)]]

par(mar = c(2, 4, 1, 1))
layout(matrix(c(1, 1, 2, 3, 3, 4), nrow = 3))
plot_cov(covs = covs_selected)

par(oldpar)

```

This gives us a general idea of what the potential viruses are.

* Plot (1) indicates that the mappability profile suggests there is areas that
can be mapped to other viruses or the host
* Plot (2) indicates that the virus segment infected the sample

## 5. References
